</> YAML
title: "Diabetes Risk Classification and Exploratory Data Analysis"
author: "Noor ul Huda"
date: "2025"
output: html_document

```{r}
library(ggplot2)
library(tidyverse)     
library(data.table) 
library(dplyr)
library(lmtest)
library(stringr)
library(magrittr)
library(randomForest)
library(caret)
library(MLmetrics)
library(smotefamily)
library(parallel)
library(pROC) 
set.seed(123)

```



```{r}
data <- read.csv("diabetes_dataset_with_notes.csv", header = TRUE, stringsAsFactors = FALSE)
View(data)
```


```{r}
summary(data)
head(data)
str(data)
```

```{r}
colSums(is.na(data))
```

```{r}
# Check for duplicate rows
sum(duplicated(data))
```

```{r}
# Remove duplicate rows
data_new <- unique(data)
```


```{r}
sum(duplicated(data_new))
```



```{r}
table(data$gender)
table(data$location)
table(data$age)
table(data$smoking_history)
table(data$diabetes)
```

```{r}
data_new <- data.table(data_new)
```


```{r}
data_new[smoking_history %in% c("ever", "not current"), smoking_history := "former"]
```

```{r}

# Remove rows where Age < 18 OR Gender is "Unknown"
data_new <- data_new %>% filter(!(age < 7 & (smoking_history == "former" | smoking_history == "current" )))

```

```{r}

# Check if any rows still have age < 7 and smoking_history as "former" or "current"
filtered_rows <- data_new %>% filter(age < 7 & (smoking_history == "former" | smoking_history == "current"))
print(filtered_rows)  # Should return 0 rows
# filtered rows (dataframe) is to make sure there are no duplicates
```

```{r}
# Check the data types of each column
sapply(data_new, class)

#convert column diabetes from binary to categorical 
data_new <- data_new %>%
  mutate(diabetes = factor(diabetes, levels = c(0, 1), labels = c("Non-Diabetic", "Diabetic")))
#delete column
# Remove column by name
data_new <- select(data_new, -clinical_notes)
```

#EDA
```{r}
ggplot(data = data_new, aes(x=diabetes, y=hbA1c_level)) + geom_boxplot() + theme_classic() + ggtitle("hbA1c vs Diabetes")

ggplot(data = data_new, aes(x=diabetes, y=blood_glucose_level)) + geom_boxplot() + theme_classic() + ggtitle("blood glucose vs Diabetes")

ggplot(data = data_new, aes(x=gender, y=hbA1c_level)) + geom_boxplot() + theme_classic() + ggtitle("hbA1c vs gender")

ggplot(data = data_new, aes(x=gender, y=bmi)) + geom_boxplot() + theme_classic() + ggtitle("bmi vs gender")

ggplot(data = data_new, aes(x=diabetes, y=bmi)) + geom_boxplot() + theme_classic() + ggtitle("bmi vs diabetes")
```

```{r}
# average blood glucose for each hbA1c level
summary_data <- data_new %>%
  group_by(diabetes) %>%
  summarise(avg_glucose = mean(blood_glucose_level, na.rm = TRUE))

# Create bar plot
ggplot(summary_data, aes(x = as.factor(diabetes), y = avg_glucose)) +
  geom_col(fill = "steelblue") +
  labs(title = "Average Blood Glucose by Diabetes",
       x = "Diabetes",
       y = "Average Blood Glucose") +
  theme_minimal()


```

```{r}
# average BMI by each HbA1c level
summary_data <- data_new %>%
  group_by(diabetes) %>%
  summarise(avg_bmi = mean(bmi, na.rm = TRUE))

# Create the bar plot
ggplot(summary_data, aes(x = as.factor(diabetes), y = avg_bmi)) +
  geom_col(fill = "darkgreen") +
  labs(title = "Average BMI by Diabetes",
       x = "Diabetes",
       y = "Average BMI") +
  theme_minimal()
```
```{r}

summary_data <- data_new %>%
  group_by(diabetes) %>%
  summarise(avg_hypertension = mean(hypertension, na.rm = TRUE))

# Create bar plot
ggplot(summary_data, aes(x = as.factor(diabetes), y = avg_hypertension)) +
  geom_col(fill = "purple") +
  labs(title = "Average Hypertension Rate by Diabetes Status",
       x = "Diabetes",
       y = "Average Hypertension Rate") +
  scale_x_discrete(labels = c("0" = "Non-Diabetic", "1" = "Diabetic")) +
  theme_minimal()

```

```{r}
cor.test(data_new$hbA1c_level , data_new$hypertension)
```

# SMOT Experiment
# Note: SMOTE was explored as a class-balancing technique,
# but the final Random Forest model below uses the cleaned original dataset.
```{r}
# Convert diabetes column to numeric directly (1 = Diabetic, 0 = Non-Diabetic)
data_df <- as.data.frame(data_new)  # Convert to data.frame if it's a data.table
data_df$diabetes <- ifelse(data_df$diabetes == "Diabetic", 1, 0)

# Select numeric columns only (excluding the target)
numeric_cols <- sapply(data_df, is.numeric)
numeric_features <- data_df[, numeric_cols]
numeric_features <- subset(numeric_features, select = -diabetes)  # Remove target from features

# Apply SMOTE
smote_output <- SMOTE(X = numeric_features, target = data_df$diabetes, K = 5)

# Build final balanced dataset
data_balanced <- smote_output$data
data_balanced$diabetes <- factor(ifelse(data_balanced$class == 1, "Diabetic", "Non-Diabetic"))

table(data_balanced$diabetes)
```

#Clustering
```{r}
#ordering smoking history
data_df$smoking_history <- as.numeric(factor(data_df$smoking_history))
#1- Current
#2- Former
#3- Never
#4- No info
```


```{r}
features <- data_df[, c("age", "bmi", "hbA1c_level", "blood_glucose_level", "smoking_history")]
```


```{r}
#Standardize the features
features_scaled <- scale(features)
```


```{r}
#K means Clustering
set.seed(42)  # For reproducibility
kmeans_result <- kmeans(features_scaled, centers = 3)
```


```{r}
data_df$cluster <- kmeans_result$cluster
```


```{r}
ggplot(data_df, aes(x = age, y = bmi, color = factor(cluster))) +
  geom_point(size = 3) +
  labs(title = "Clusters of Diabetes Patients", x = "Age", y = "BMI")
```


```{r}
# Determine optimal number of clusters using the Elbow method
wss <- sapply(1:10, function(k) { kmeans(features_scaled, centers = k, nstart = 25)$tot.withinss })
plot(1:10, wss, type = "b", pch = 19, frame = FALSE, xlab = "Number of Clusters (k)", ylab = "Total Within-Cluster Sum of Squares")
 
# Apply K-means with chosen k (e.g., k = 3)
set.seed(123)  # For reproducibility
k <- 4  # Choose based on the elbow plot
kmeans_result <- kmeans(features_scaled, centers = k, nstart = 25)
 
# Add cluster labels to the original dataset
data_df$cluster <- as.factor(kmeans_result$cluster)
 
# Print cluster summary
print(table(data_df$cluster))

 
# Visualizing clustering results (if reducing dimensions)
library(ggplot2)
data_df$Cluster <- as.factor(kmeans_result$cluster)
ggplot(data_df, aes(x = bmi, y = blood_glucose_level, color = Cluster)) +
  geom_point(size = 3) + theme_minimal() +
  labs(title = "K-means Clustering of Features", x = "BMI", y = "Blood Glucose Level")
```


# Machine Learning Pediction
```{r}
start_time <- Sys.time()

# Split data
index <- createDataPartition(data_new$diabetes, p = 0.8, list = FALSE)
train_data <- data_new[index, ]
test_data <- data_new[-index, ]

rf_model <- randomForest(
  diabetes ~ ., 
  data = train_data,
  ntree = 500,             
  mtry = 4,
  importance = TRUE,
  nodesize = 5,
  maxnodes = 30
)

# Predict
rf_predictions <- predict(rf_model, newdata = test_data)


```

```{r}
# Confusion matrix
conf_matrix <- confusionMatrix(rf_predictions, test_data$diabetes)
print(conf_matrix)

# Extract values
cm <- conf_matrix$table
TP <- cm["Diabetic", "Diabetic"]
FN <- cm["Non-Diabetic", "Diabetic"]
FP <- cm["Diabetic", "Non-Diabetic"]
TN <- cm["Non-Diabetic", "Non-Diabetic"]

# Calculate recall and F1 score manually
recall <- TP / (TP + FN)
precision <- TP / (TP + FP)
f1_score <- 2 * ((precision * recall) / (precision + recall))

# Print them
cat("Recall:", round(recall, 3), "\n")
cat("F1 Score:", round(f1_score, 3), "\n")

# ROC curve
# For ROC, you need probabilities, not classes
rf_probabilities <- predict(rf_model, newdata = test_data, type = "prob")

# Plot ROC
roc_obj <- roc(response = test_data$diabetes, predictor = rf_probabilities[, "Diabetic"])
plot(roc_obj, main = "ROC Curve - Random Forest", col = "lightblue")
auc_value <- auc(roc_obj)
cat("AUC:", round(auc_value, 3), "\n")

# End time
end_time <- Sys.time()

# Total time taken
cat("Time taken:", round(difftime(end_time, start_time, units = "secs"), 2), "seconds\n")
```

#Importance Graph
```{r}
# Plot top important variables
varImpPlot(rf_model,
           type = 2,
           main = "Feature Importance",
           col = "blue",
           cex.main = 1.2,
           cex.axis = 0.8)

importance(rf_model)

```


# SparkR Distributed Processing
# This section explores distributed data processing with SparkR.
# It applies rule-based classification thresholds rather than training
# the Random Forest model used in the previous section.
```{r}
Sys.setenv(JAVA_HOME = "/Library/Java/JavaVirtualMachines/jdk-17.jdk/Contents/Home")

```

```{r}

Sys.setenv(SPARK_HOME = "/Users/ammarkhan/Desktop/spark-3.5.5-bin-hadoop3") 

.libPaths(c(file.path(Sys.getenv("SPARK_HOME"), "R", "lib"), .libPaths()))

start_time <- Sys.time()

# Load SparkR
library(SparkR)

# Start Spark session
sparkR.session(master = "local[*]", appName = "DiabetesRiskClassification")
```

```{r}
# Load your cleaned dataset into a Spark DataFrame
df <- as.DataFrame(data_new)

with_diabetes_prediction <- withColumn(
  df,
  "predicted_diabetes",
  when(
    df$hbA1c_level > 6.5 | df$blood_glucose_level > 140 | df$bmi > 30, "Diabetic"
  ) %>%
    otherwise("Non-Diabetic")
)

# Aggregate confusion matrix
confusion_matrix <- summarize(
  groupBy(with_diabetes_prediction, with_diabetes_prediction$diabetes, with_diabetes_prediction$predicted_diabetes),
  count = n(with_diabetes_prediction$diabetes)
)
final_confusion <- collect(confusion_matrix)
print(final_confusion)

```

```{r}
# Pivot table: rows = actual, columns = predicted
confusion_table <- final_confusion %>%
  tidyr::pivot_wider(names_from = predicted_diabetes, values_from = count, values_fill = 0)

# Assign values
TP <- confusion_table$Diabetic[confusion_table$diabetes == "Diabetic"]
FN <- confusion_table$`Non-Diabetic`[confusion_table$diabetes == "Diabetic"]
FP <- confusion_table$Diabetic[confusion_table$diabetes == "Non-Diabetic"]
TN <- confusion_table$`Non-Diabetic`[confusion_table$diabetes == "Non-Diabetic"]

# Calculate metrics
accuracy <- (TP + TN) / (TP + TN + FP + FN)
recall <- TP / (TP + FN)
precision <- TP / (TP + FP)
f1_score <- 2 * (precision * recall) / (precision + recall)

# Print metrics
cat("Accuracy:", round(accuracy, 3), "\n")
cat("Recall:", round(recall, 3), "\n")
cat("F1 Score:", round(f1_score, 3), "\n")

# Optional: Visualise
barplot(height = c(TP, FP, FN, TN),
        names.arg = c("TP", "FP", "FN", "TN"),
        col = c("violet", "red", "black", "blue"),
        main = "Confusion Matrix Components",
        ylab = "Count")

# Stop Spark session
sparkR.session.stop()

# End time
end_time <- Sys.time()

# Total time taken
cat("Time taken:", round(difftime(end_time, start_time, units = "secs"), 2), "seconds\n")

```



